---
title: "ggtree with binary data"
author: "Albert Vill"
date: "2023-07-26"
output:
  pdf_document: default
  html_document: default
---

# load packages
```{r, message=FALSE, warning=FALSE}
# requires native pipe (R version >= 4.1)
library(tidyverse) # v3.3.6
library(ggtree) # v3.4.2
library(ggtreeExtra) # v1.6.1
library(ape) # v5.6.2
library(phytools) # v1.0.3
library(patchwork) # v1.1.2
library(ggnewscale) # v0.4.8
```

# MOCK DATA

## make an `m × n` binary matrix
```{r}
binmat <- function(m, n, prob) {
  mat <- matrix(data = NA, nrow = m, ncol = n)
  for (i in 1:m) {
    mat[i,] <- rbinom(n = n, size = 1, prob = prob)
  }
  rownames(mat) <- paste0("N", sprintf(paste0("%0",nchar(m),"d"), 1:m))
  mat
}
```

## generate mock tree
```{r}
set.seed(123)
tree <- binmat(m = 25, n = 50, prob = 0.7) |>
  dist.gene(method = "pairwise") |>
  nj()
```

## visualize rectangular tree
```{r}
ggtree(tree, layout = "rectangular") +
  geom_tiplab(size = 4,
              align = T,
              linetype = "dotted",
              linesize = 0.7,
              offset = 1,
              hjust = -0.1)
```

## color branches by clade membership
```{r}
# select clades by most recent common ancestor
node1 <- findMRCA(tree, tips = c("N24","N04"))
node2 <- findMRCA(tree, tips = c("N05","N16"))
node3 <- findMRCA(tree, tips = c("N17","N13"))

tree1 <- groupClade(tree,
                    .node = c(node1, node2, node3),
                    group_name = "clades")

pal <- c("#000000","#00A08A","#F2AD00","#FF0000")

ptree <- ggtree(tr = tree1,
       mapping = aes(color = clades),
       layout = "rectangular") +
  geom_tiplab(size = 4,
              align = T,
              linetype = "dotted",
              linesize = 0.7,
              offset = 1,
              hjust = -0.1) +
  scale_color_manual(values = pal) +
  geom_cladelabel(node = node1, label = "Clade X", 
                  align = T,
                  fontsize = 5,
                  offset = 1.8,
                  color = pal[2]) +
  geom_cladelabel(node = node2, label = "Clade Y",
                  align = T,
                  fontsize = 5,
                  offset = 1.8,
                  color = pal[3]) +
  geom_cladelabel(node = node3, label = "Clade Z",
                  align = T,
                  fontsize = 5,
                  offset = 1.8,
                  color = pal[4]) +
  theme(legend.position = "none",
        plot.margin = margin(0,30,0,0,"pt"))

ptree + ggplot2::coord_cartesian(clip = "off")
```

# PHAM DATA

## read in pham table and subset for non-orphams
```{r}
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022/")
#library(tidyverse)

phamdat <- read_csv(file = "pham_table.txt",
                    col_names = c("name", "acc", paste0("pham_",1:10312)))
phamdat_noOrphams <- phamdat |> select(!where( ~ is.numeric(.x) && sum(.x) <= 1))
```

### which phams are unique to W3 phages
```{r}
setwd("C:/Users/acvill/Dropbox/Phage/BacillusPaper-150Kb/J1_PhiNIT1/ani_aai_matrix/")
acc2name <- readr::read_tsv(file = "acc2name.txt",
                            col_names = c("seqname", "phage"))
'%ni%' <- Negate("%in%")
W3_phams <- phamdat |>
  select(-name) |>
  pivot_longer(cols = -acc) |>
  filter(acc %ni% acc2name$seqname) |>
  group_by(name) |>
  summarise(sum = sum(value)) |>
  filter(sum == 0) |>
  select(name) |> as_vector()

W3_phage2phams <- phamdat |>
  select(-name) |>
  pivot_longer(cols = -acc) |>
  filter(name %in% W3_phams, acc %in% acc2name$seqname, value > 0) |>
  left_join(rename(acc2name, acc = seqname)) |>
  group_by(name) |>
  summarize(sum = n(), phages = toString(sort(phage))) |>
  rename(pham = name)
```

### create pham2protein and phage2prot mapping tables
```{bash}
cd /c/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022
tar -xzvf pham_fastas.tar.gz
cd pham_fastas
while read fasta; do
  pham=$(echo "${fasta}" | sed "s/.faa//g")
  grep ">" ${fasta} | \
  sed "s/^>/${pham}\t/g" | \
  sed "s/_prot_/\t/g"
done < <(ls | grep ".faa") > pham2protein.txt
mv pham2protein.txt ../
cd ..
grep ">" Bacillus_355_db.faa | \
  sed "s/>lcl|//g" | \
  sed "s/\] \[/\t/g" | \
  sed "s/ \[/\t/g" | \
  sed "s/]$//g" | \
  sed "s/_prot_/\t/g" > phage2prot2func.txt
```

### merge pham, protein, phage data
```{r}
setwd("C:/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022/")
pham2prot <- read_tsv(file = "pham2protein.txt",
                      col_names = c("pham","acc","protein"))
phage2prot <- read_tsv(file = "phage2prot2func.txt",
                       col_names = c("acc", "protein", "gene",
                                     "locus", "function", "prot_id"))

pham2function <- phage2prot |>
  filter(acc %in% acc2name$seqname) |>
  mutate(Function = ifelse(grepl("protein_id=", `function`), locus, `function`)) |>
  select(acc, protein, Function) |>
  mutate(Function = str_remove(Function, "protein=")) |>
  left_join(pham2prot, by = c("acc", "protein")) |>
  left_join(rename(acc2name, acc = seqname), by = "acc") |>
  select(pham, Function) |> group_by(pham) |>
  summarize(func = toString(sort(unique(Function)))) |>
  ungroup()

W3_pham2func <- left_join(W3_phage2phams, pham2function, by = "pham")
write_tsv(x = W3_pham2func,
          file = "C:/Users/acvill/Dropbox/Phage/BacillusPaper-150Kb/J1_PhiNIT1/W3_unique_phams_functions.tsv")
```

### assign function to W3 unique phams

## convert data to named matrix
```{r}
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022/")
toMat <- function(data) {
  
  rn <- data |> select(name) |> as_vector()
  cn <- data |> select(-c(name,acc)) |> colnames()
  mat <- data |> select(-c(name,acc)) |> as.matrix()
  rownames(mat) <- rn
  colnames(mat) <- cn
  mat
  
}

pmat_all <- toMat(phamdat)
pmat_noOrphams <- toMat(phamdat_noOrphams)
```

## build nj tree
```{r}
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022/")
tree_all <- nj(dist.gene(x = pmat_all,method = "pairwise"))
tree_noOrphams <- nj(dist.gene(x = pmat_noOrphams, method = "pairwise"))
```

## visualize tree
```{r}
rec_tree <- function(tree, tiplabs) {

  pal <- c("#0000FF","#FF0000")
  g2_node <- findMRCA(tree, tips = tiplabs)
  tree1 <- groupClade(tree,
                      .node = g2_node,
                      group_name = "clades")
  
  pt <- ggtree(tree1,
               layout = "rectangular",
               branch.length = "none",
               ladderize = T,
               right = T,
               mapping = aes(color = clades)) +
    scale_color_manual(values = pal) +
    theme(legend.position = 'none')
  
  pu <- ggtree(tree1,
               layout = "rectangular",
               branch.length = "none",
               ladderize = T,
               right = T,
               mapping = aes(color = clades)) +
    scale_color_manual(values = pal) +
    theme(legend.position = 'none')
  
  # (pt + (viewClade(tree_view = pt,
  #                  node = MRCA(pt, tiplabs),
  #                  xmax_adjust = 15) +
  #         geom_tiplab() +
  #         theme(legend.position = 'none'))) &
  #   theme(panel.border = element_rect(colour = "black",
  #                                     size = 1,
  #                                     fill = NA))

  # pt + inset_element(viewClade(tree_view = pt,
  #                              node = MRCA(pt, tiplabs),
  #                              xmax_adjust = 15) +
  #                      geom_tiplab() +
  #                      theme(legend.position = 'none') +
  #                      theme(panel.border = element_rect(colour = "black",
  #                                                        size = 1,
  #                                                        fill = NA)),
  #                    left = 0,
  #                    bottom = 0,
  #                    right = 0.4,
  #                    top = 0.4)
  
  open_tree(pt,90) + inset_element(viewClade(tree_view = pu,
                               node = MRCA(pt, tiplabs),
                               xmax_adjust = 15) +
                       geom_tiplab() +
                       theme(legend.position = 'none') +
                       theme(panel.border = element_rect(colour = "black",
                                                         size = 2,
                                                         fill = NA)),
                     left = 0.55,
                     bottom = 0.10,
                     right = 0.9,
                     top = 0.45)
  
}

g2 <- c("019DV002","019DV004","056SW001B","276BB001",
        "280BB001","274BB002","268TH004")
print_all <- rec_tree(tree_all, g2)
print_sub <- rec_tree(tree_noOrphams, g2)

print_all
print_sub
```

### ggsave
```{r}
#library(svglite)
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_14Oct2022/trees/")

# include shape, branch length, pham set, version
ggsave(filename = "BacillusPhageTree_circ_allPhams_noBranchLength_v3.pdf",
       plot = print_all, width = 10, height = 10, units = "in")
ggsave(filename = "BacillusPhageTree_circ_noOrphams_noBranchLength_v3.pdf",
       plot = print_sub, width = 10, height = 10, units = "in")
```

## read in bacphlip data and get phage names from pham data
```{r}
# Virulent = 1 - Temperate
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-52Kb/phammseqs_17Oct2022/")
bacdat <- read_tsv(file = "Bacillus_355_db.fna.bacphlip", col_names = T) |>
  rename(acc = ...1, virulent = Virulent) |> select(acc, virulent) |>
  left_join(phamdat |> select(name, acc), by = "acc")

# check distribution
bacdat |> select(virulent) |> as_vector() |> hist(breaks = 20)

# enframe Geobacillus squares
gb_names <- c("GBK1","GBK2","GBSV1","TP-84","GVE2","GVE3")

gb_nodedat <- sapply(gb_names, function(x)
  {which(tree_noOrphams$tip.label == x)
  }) |>
  enframe() |> rename(node = value) |> mutate(frame = "geo")

bacdat <- bacdat |> left_join(gb_nodedat) |> mutate(frame = ifelse(is.na(frame), "not_geo", "geo"))
```

## create a tree with bacphlip annotation
```{r}
anno_tree <- function(tree, highlight, ringdat, tiplist) {

  pal <- c("#808080","#FF0000","#0000FF")
  
  pt <- ggtree(tree,
               layout = "rectangular",
               branch.length = "none",
               ladderize = T,
               right = T,
               mapping = aes(color = clades)) +
    scale_color_manual(values = pal) +
    geom_tiplab2(align = TRUE,
                  linetype = 3,
                   size = 1,
                   linesize = 0.2,
                   show.legend = FALSE,
                   offset = 2) +
    ggnewscale::new_scale_color() +
    geom_fruit(data = ringdat,
               geom = geom_tile,
               mapping = aes(y = name, fill = virulent, color = frame),
               width = 0.8, height = 0.8) +
    scale_color_discrete(type = c("black","white")) +
    scale_fill_gradient2(high = "#ef8a62", low = "#67a9cf",
                         mid = "#f7f7f7", midpoint = 0.5,
                         guide = guide_colorbar(label = TRUE,
                                              frame.colour = "black",
                                              ticks = TRUE,
                                              nbin = 100,
                                              label.position = "bottom",
                                              ticks.colour = "black",
                                              #barwidth = 13,
                                              barheight = 1.3,
                                              direction = 'horizontal'))
    # geom_fruit(data = tipdat,
    #            geom = geom_point,
    #            mapping = aes(y = name),
    #            size = 1.5,
    #            show.legend = FALSE,
    #            pch = 20,
    #            color = "black")
    
  
  pu <- ggtree(tree,
               layout = "rectangular",
               branch.length = "none",
               ladderize = T,
               right = T,
               mapping = aes(color = clades)) +
    scale_color_manual(values = pal) +
    theme(legend.position = 'none')
  
  # (pt + (viewClade(tree_view = pt,
  #                  node = MRCA(pt, tiplabs),
  #                  xmax_adjust = 15) +
  #         geom_tiplab() +
  #         theme(legend.position = 'none'))) &
  #   theme(panel.border = element_rect(colour = "black",
  #                                     size = 1,
  #                                     fill = NA))

  # pt + inset_element(viewClade(tree_view = pt,
  #                              node = MRCA(pt, tiplabs),
  #                              xmax_adjust = 15) +
  #                      geom_tiplab() +
  #                      theme(legend.position = 'none') +
  #                      theme(panel.border = element_rect(colour = "black",
  #                                                        size = 1,
  #                                                        fill = NA)),
  #                    left = 0,
  #                    bottom = 0,
  #                    right = 0.4,
  #                    top = 0.4)
  
  open_tree(pt,90) +
    patchwork::inset_element(
      p = ggtree::viewClade(tree_view = pu,
                            node = ggtree::MRCA(pu, highlight),
                            xmax_adjust = 30) +
        geom_tiplab() +
        theme(legend.position = 'none') +
        theme(panel.border = element_rect(colour = "black",
                                          size = 2,
                                          fill = NA)),
      left = 0.55,
      bottom = 0.10,
      right = 0.9,
      top = 0.45)
  
}

# add clade annotations to tree before plotting

# g2_node <- findMRCA(tree_noOrphams, tips = c("019DV002","019DV004","056SW001B","276BB001",
#                                              "280BB001","274BB002","268TH004"))

# j1_node_extra  <- findMRCA(tree_noOrphams,
#                            tips = c("010DV004","010DV005","035JT004","043JT007",
#                                     "278BB001","phiNIT1_DNA", "Grass", "BSP10",
#                                     "BSP9", "BSTP3"))

j1_node <- findMRCA(tree_noOrphams, tips = c("010DV004","010DV005","035JT004",
                                             "043JT007", "278BB001"))

tree_noOrphams2 <- groupClade(tree_noOrphams,
                              .node = c(j1_node),
                              group_name = "clades")

tohilite <- findMRCA(tree_noOrphams2, tips = c("010DV004","010DV005","035JT004",
                                             "043JT007", "278BB001"))

print_annot <- anno_tree(tree = tree_noOrphams2,
                         highlight = tohilite,
                         ringdat = bacdat,
                         tiplist = c("010DV004","010DV005","035JT004",
                                             "043JT007", "278BB001"))

print_annot
```

## ggsave
```{r}
#library(svglite)
setwd("/Users/acvill/Dropbox/Phage/BacillusPaper-150Kb/J1_PhiNIT1/tree_hgt/")

# include shape, branch length, pham set, version
ggsave(filename = "BacillusPhageTree_circ_noOrphams_noBranchLength_v2.pdf",
       device = cairo_pdf,
       plot = print_annot, width = 10, height = 10, units = "in")
```